A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2022-11-23 pilot-publication/manuscript.Rmd
Complete preliminary discussion

## Data monitoring
Weekly meetings with the core team and project officers will take
place and for this meeting a data status report will be automatically
generated highlighting missing data and number of patients awaiting
followup. Cluster specific interim analysis will take place after two
months. The results of this will be presented to the core team, this
team will decide if the pilot should be terminated. Although we will
not have formal termination criteria because of the short duration of
the study, reasons not to continue could include that the collection
of key variables, such as mortality outcomes, is unfeasible or that
patients are not consenting to be included in the data
collection. A data monitoring committee
will not be used in the pilot study due to its limited scope.

@@

Previous studies on the effect of in-hospital trauma life support
training on patient outcomes are observational or quasi-experimental
without a control group, with heterogeneous results
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
@Wang2010]. Studies from Trinidad and Tobago and, El Salvador, Rwanda,
and Cambodia found no significant effect on patient mortality after
implementing in-hospital trauma life support training
[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
whereas one study from China on 820 patients found a significant
reduction in mortality, from 20 to 15%, after implementing ATLS
[@Wang2010].

Several randomised controlled trials show that ATLS is associated with
improved provider skills.

@@

Compare missing data rates.

Non-inferiority vs superiority.

@@

>>>>

## Data monitoring
Weekly meetings with the core team and project officers will take
place and for this meeting a data status report will be automatically
generated highlighting missing data and number of patients awaiting
followup. We conducted an interim analysis on October 12, 2022, and
decided to complete the study as participants were consenting to be
included in the study and key variables including mortality outcomes
could be collected. No outcomes were dropped. We did not use a data
monitoring committee in the pilot study due to its limited scope.

@@

@@

<!-- Consider discussing non-inferiority vs superiority. -->

During the course of this pilot we deviated from the protocol in
multiple ways, and provide a detailed list as Supplementary
material. Some key limitations of this pilot and therefore lessons to
be learned and factored into the design of the full-scale trial
include the lower than expected enrolment rates of some centres,
centre specific management routines, and difficulties in collecting
data on complications and cause of death. We minimised the impact of
the lower than expected enrolment rates by including a seventh centre,
but on-site observations of patient volumes are likely to be needed
for the full-scale trial. We decided to be pragmatic in selecting
which residents to train and how to structure the data collection
depending on how and by whom patients were initially managed, but this
flexibility will need to be built into the full-scale trial
protocol. Finally, we found that data on complications and cause of
death were hard to identify and therefore the full-scale trial will
need to include longer training of project officers.

Previous studies on the effect of in-hospital trauma life support
training on patient outcomes are observational or quasi-experimental
without a control group, with heterogeneous results
[@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; VanHeng2008;
@Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and
Cambodia found no significant effect on patient mortality after
implementing in-hospital trauma life support training
[@Ariyanayagam1992; @Cioèpeña2016; @Petroze2014; @Vanheng2008],
whereas one study from China that included 820 patients found a
significant reduction in mortality, from 20 to 15%, after implementing
ATLS [@Wang2010].

Several controlled trials, including at least two randomized
controlled trials [@Ali1996; @Ali1999], show that ATLS is associated
with improved provider skills [@Mohammad2013]. Observational evidence
indicates that PTC also leads to improved provider skills
[@Kadhum2020]. The missing link is then how, and if, these improved
skills translate into improved patient outcomes. As trauma care
providers we assume, and probably rightly so, that we deliver better
care if we train. The question is then how we should train, especially
considering the costs associated with some of the programs
offered.

We conclude that a full-scale cluster randomised trial should be
feasible after incorporating the lessons of this pilot, and that this
full-scale trial should, regardless of its outcome, influence how we
train trauma care providers in the future.

@@

## Data collection from records
We decided to record data from records only for a subset of patients
to reduce the project officers' workload.
